\systemEnding: Bootstrapping Domain-Specific Languages for Neurally-Guided Bayesian Program Learning

### 2018-06-13 archive.tex
progress on archive posting

The FlashFill system~\cite{gulwani2011automating} embedded in
Microsoft Excel solves problems such as these and is probably the best
known practical program-induction algorithm, but researchers in
programming languages and AI have had successes in many domains, such as handwriting recognition and
generation~\cite{lake2015human}, procedural
graphics~\cite{ellis2017learning}, question
answering~\cite{johnson2017clevr} and robot motion
planning~\cite{devlin2017neural}.  These systems
work in different ways, but most hinge upon a carefully engineered
\textbf{Domain Specific Language (DSL)}.  This is especially true for
systems like FlashFill that induce a wide range of programs
very quickly, in a few seconds or less.  DSLs constrain the search
over programs with strong prior knowledge in the form of a restricted
set of programming primitives finely tuned to the domain: for
text editing, these are operations like appending and
splitting on characters.

In this work, we consider the problem of building agents that learn to solve
program induction tasks, and also the problem of acquiring the prior knowledge
necessary to quickly solve these tasks in a new domain.  Representative problems
in three domains are shown in Table~\ref{initialExampleDSL}.  Our solution is an
algorithm that grows or boostraps a DSL while jointly training a neural network
to help write programs in the increasingly rich DSL.
%
% (JBT: MAYBE INSERT FIGURE FROM ICML PAPER?)
%
\begin{table*} %[t!]
  \makebox[\textwidth][c]{
    \scriptsize
  \tabcolsep=4pt
  \renewcommand\code\texttt
  \renewcommand\codechar[1]{\texttt{"#1"}}
  \newcommand{\helpSize}{0.25cm}
  \begin{tabular}{>{\hspace{-0em}}c<{\hspace{-1em}}>{\hspace{-1em}}c<{\hspace{-1em}}>{\hspace{-2.5em}}c<{\hspace{-0.8em}}>{\hspace{-1em}}c<{\hspace{-1em}}}
    \toprule
    &{\normalsize List Functions}&{\normalsize Text Editing}&{\normalsize Symbolic Regression}\\\midrule
    \rotatebox[origin=c]{90}{\normalsize \pop{Programs} \& Tasks}&{\tabcolsep=7pt
      \begin{tabular}{cc}
        \begin{tabular}{c}
          \code{[7\, 2\, 3]}$\to $\code{[7\, 3]}         \\
          \code{[1\, 2\, 3\, 4]}$\to $\code{[3\, 4]} \\
          \code{[4\, 3\, 2\, 1]}$\to $\code{[4\, 3]} \\
          \pop{\code{$f(\ell) = $}\code{($f_1$ $\ell$ ($\lambda$ (x)}}\\
          \hspace{1.15cm}\pop{\code{(> x 2)))}}       \\
          \\
          \code{[2\, 7\, 8\, 1]}$\to $\code{8}               \\
          \hspace{0.15cm}\code{[3\, 19\, 14]}$\to $\code{19}                \\
          \pop{\code{$f(\ell) = $}\code{($f_2$ $\ell$)}}
        \end{tabular}
        &
        \hspace{-0.3cm}\begin{tabular}{c}
          \code{[7\, 3]}$\to $\code{False}                              \\
          \hspace{0.3cm}\code{[3]}$\to $\code{False}                    \\
          \hspace{-0.3cm}\code{[9\, 0\, 0]}$\to $\code{True\phantom{e}} \\
          \hspace{0.3cm}\code{[0]}$\to $\code{True\phantom{e}}                        \\
          \hspace{-0.3cm}\code{[0\, 7\, 3]}$\to $\code{True\phantom{e}}                \\
          \pop{\code{$f(\ell) = $}\code{($f_3$ $\ell$ 0)}}
        \end{tabular}
      \end{tabular}
    }
    &
    \hspace{-0.5cm}\begin{tabular}{c}
      +106 769-438$\to $106.769.438\\%&Nancy FreeHafer $\longrightarrow$ Dr. Nancy\\
      +83 973-831$\to $83.973.831\\
      \pop{$f(\text{\code{s}}) = $\code{(}$f_0$\code{  \codechar{.} \codechar{-}
      }}\\
      \hspace{1.25cm}\pop{\code{($f_0$ \codechar{.} \codechar{ }}}\\
      \hspace{1.4cm}\pop{\code{(cdr s)))}}\\
      ~\\
      Temple Anna H $\to $TAH\\
      Lara Gregori$\to $LG\\
      \pop{$f(\text{\code{s}}) = $\code{(}$f_2$\code{ s)}}\\
    \end{tabular}
    &
    \begin{tabular}{cc}
      \includegraphics[width = 3em]{figures/functions/4.png}&
      \includegraphics[width = 3em]{figures/functions/146}\\
      \pop{\code{$f($x$) = $($f_1$ x)}}&    \pop{\code{$f($x$) = $($f_6$ x)}}\\
      ~\\
      \includegraphics[width = 3em]{figures/functions/112.png}&
        \includegraphics[width = 3em]{figures/functions/92.png}
      \\
      \pop{\code{$f($x$) = $($f_4$ x)}}&    \pop{\code{$f($x$) = $($f_3$ x)}}\\

@@

  \end{tabular}
  \\\bottomrule\\
\end{tabular}}\vspace{-0.5cm}
\caption{Top: Tasks from three domains we apply our algorithm to, each followed by the programs \system discovers for them. Bottom: Several examples from learned DSL. Notice that learned DSL primitives can call each other, and that \system rediscovers higher-order functions like \code{filter} ($f_1$ under List Functions)}\label{initialExampleDSL}\vspace{-0.5cm}
\end{table*}
Because any learning problem can in principle be cast as
program induction, it is important to delimit our focus.  In contrast
to computer assisted programming~\cite{solar2008program} or genetic
programming~\cite{DBLP:books/daglib/0070933}, our goal is not to automate software
engineering, or to synthesize large bodies of code starting from scratch.  Ours is a basic AI goal:
capturing the human ability to learn to think flexibly and efficiently
in new domains --- to learn what you need to know about a domain so you
don't have to solve new problems starting from scratch.  We are
focused on problems that people solve relatively
quickly, once they acquire the relevant domain expertise.  These
correspond to tasks solved by short programs --- if you have an
expressive DSL.  %% Even with a good DSL, program search may be
%% intractable;
%
% , and adding new library routines to the DSL only broadens the
% search space;
%
%% so we train a
%% neural network to guide the search procedure.

@@

We take inspiration from several ways that skilled human
programmers have learned to code: Skilled coders build libraries of
reusable subroutines that are shared across related programming tasks,
and can be composed to generate increasingly complex
subroutines.  In text editing, a good library should support routines
for splitting on characters, but also specialize these routines to
split on frequent delimiters such as spaces or commas. Skilled coders
also learn to recognize what kinds of programming idioms and library
routines would be useful for solving the task at hand, even if they
cannot instantly work out the details.  In text editing, one might
learn that if outputs are consistently shorter than inputs, removing
characters is likely part of the solution.%% ; if every output
%% contains a constant substring (e.g., ``Dr.''), inserting or appending
%% that constant string is likely to be a subroutine.

Our algorithm is called \system because it is based on a
novel kind of ``wake-sleep'' learning (c.f. \cite{hinton1995wake}), iterating
between ``wake'' and ``sleep'' phases to achieve three goals: finding
programs that solve tasks; creating a DSL by discovering
and reusing domain-specific subroutines; and training a neural network
that efficiently guides search for programs in the DSL.
The learned DSL effectively encodes a prior on programs likely to
solve tasks in the domain, while the neural net looks at the example
input-output pairs for a specific task and produces a ``posterior''
for programs likely to solve that specific task.  The neural network
thus functions as a \textbf{recognition model} supporting a form of
approximate Bayesian program induction, jointly trained with a
\textbf{generative model} for programs encoded in the DSL, in the
spirit of the Helmholtz machine~\cite{hinton1995wake}. The
recognition model ensures that searching for programs remains
tractable even as the DSL %% (and hence the search space for programs)
expands.

%% Our \system
%% algorithm incorporates these insights by iterating
Concretely, our algorithm iterates through three steps. 
The \textbf{Wake} step takes a given set of \textbf{tasks}, typically
several hundred, and searches for compact programs that solve these tasks,
guided by the current DSL and neural network.
The \textbf{Sleep-G} step grows the DSL (or \textbf{G}enerative model) 
which allows the agent to more compactly
write programs in the domain; it modifies the structure of the DSL by
discovering regularities across programs found during waking, compressing
them to distill out common code fragments across successful programs.
The \textbf{Sleep-R} step improves the search procedure by training a neural network (the \textbf{R}ecognition model) to
write programs in the current DSL, in the spirit of ``amortized'' or
``compiled'' inference~\cite{le2016inference}. %%  (CITE ,
%% and others too on amortized inference.)
%Crucially, these steps bootstrap off each other~(Figure~\ref{feeding}).
%% \begin{figure}[h]\centering
%%   \begin{tikzpicture}
%%     \begin{scope}[shift = {(1,-1)}]
%%     \node[align = center](synthesis) at (6,4) {Search for \\programs};
%%     \node[align = center](DSL) at (3,1) {DSL};
%%     \node[align = center](recognitionModel) at (9,1) {Neural\\recognition \\model};

%%     \draw [->,thick] (synthesis.-120) to[out = -150,in = 60] node[below,rotate = 45,align = center]{{\footnotesize Trains}\\{\footnotesize (Sleep-G)}} (DSL.30);
%%     \draw [->,thick] (synthesis.-60) to[out = -30,in = 120] node[below,rotate=-45,align = center]{{\footnotesize Trains}\\{\footnotesize (Sleep-R)}} (recognitionModel.150);
%%     \draw [->,thick] (DSL.east) to[out = -30,in = 210] node[above, align = center]{{\footnotesize Trains}\\{\footnotesize (Sleep-R)}} (recognitionModel.west);

%%     \draw [->,thick,dashed] (DSL.north) to[out = 90,in = 180] node[fill=white,align = center]{  \footnotesize{Inductive bias}\\\footnotesize{(Wake)}} (synthesis.west);
%%     \draw [->,thick,dashed] (recognitionModel.north) to[out = 90,in = 0] node[fill=white,align = center]{{\footnotesize Makes tractable}\\{\footnotesize (Wake)}} (synthesis.east);
%%   \end{scope}
%%     \end{tikzpicture}
%%   \caption{Algorithm solves for programs, the DSL, and a recognition model. These steps bootstrap off each other in a Helmholtz-machine inspired wake/sleep inference strategy.}  \label{feeding}
%% \end{figure}

%% We apply \system to three domains:
%% list processing; text editing (in the style of FlashFill~\cite{gulwani2011automating}); and symbolic regression.
%%  For each of these we initially provide a generic
%%  set of programming primitives.
%% Our algorithm then constructs
%% its own DSL for expressing solutions in the domain (Tbl.~\ref{initialExampleDSL}).

\textbf{Related work.} Prior work on program learning has largely assumed a fixed, hand-engineered DSL,
both in classic symbolic program learning approaches (e.g., Metagol:~\cite{muggleton2015meta},
FlashFill:~\cite{gulwani2011automating}),
neural approaches  (e.g., RobustFill:~\cite{devlin2017robustfill}), and hybrids of neural and
symbolic methods (e.g., Neural-guided deductive search:~\cite{ngds}, DeepCoder:~\cite{balog2016deepcoder}).
A notable exception is the EC algorithm~\cite{Dechter:2013:BLV:2540128.2540316},
which also learns a library of subroutines.
We find EC motivating, and go beyond it and other prior work through the following contributions: %% \begin{itemize}
%%   \setlength\itemsep{0}
%%   \item We show how to learn-to-learn programs in an expressive Lisp-like programming language;
%%   \item We give an algorithm for learning DSLs, built on a formalism known
%%     as Fragment Grammars~\cite{tim}; and 
%%   \item We give a hierarchical Bayesian framing of the problem that allows a
%%     joint inference of the DSL and recognition model.
%% \end{itemize}
\\\textbf{Contributions.} (1) We show how to learn-to-learn programs in an expressive Lisp-like programming language, including conditionals, variables, and higher-order recursive functions;
(2) We give an algorithm for learning DSLs,
built on a formalism known as Fragment Grammars~\cite{tim};
and (3) We give a hierarchical Bayesian framing enabling joint inference of the DSL and recognition model.

@@

\system takes as input a set of \emph{tasks}, written $X$, each of which is a program synthesis problem.
It has at its disposal a domain-specific \emph{likelihood model}, written $\probability[x|p]$, which scores the likelihood of a task $x\in X$ given a program $p$. %% \footnote{For example, for string editing,
%%   the likelihood is 1 if the program predicts the observed outputs on the observed inputs,
%% and 0 otherwise.}
Its goal is to solve each of the tasks by writing a program,
and also to infer a DSL, written $\mathcal{D}$,
which is a generative model over programs, written $\probability[p |\mathcal{D}]$.
We frame our goal as maximum a posteriori (MAP) inference of $\mathcal{D}$ given $X$:
$$
\argmax_{\mathcal{D}}\probability[\mathcal{D}]\prod_{x\in X}\sum_p\probability[x|p]\probability[p|\mathcal{D}]
$$
Even evaluating the above objective is intractable due to the sum over all programs,
and so we instead marginalize over finite sets of programs, called \textbf{frontiers}, one for each task $x\in X$, written $\mathcal{F}_x$, giving the following lower bound on the marginal, which we call $\lowerBound$:
\begin{equation}\label{lowerBound}
  \lowerBound  \triangleq\probability[\mathcal{D}]\prod_{x\in X}\sum_{p\in \mathcal{F}_x}\probability[x|p]\probability[p|\mathcal{D}]
\end{equation}
We alternate maximization of $\lowerBound$ w.r.t. $\left\{\mathcal{F}_x \right\}_{x\in X}$ %% (\textbf{Wake})
and $\mathcal{D}$:  %% (\textbf{Sleep-G})

\\\noindent \textbf{Wake: Maxing $\lowerBound$ w.r.t.\ the frontiers.} Here $\mathcal{D}$ is fixed and we
want to find new programs to add to  the frontiers so that $\lowerBound$ increases the most.
$\lowerBound$ most increases by finding programs where $\probability[x|p]\probability[p|\mathcal{D}]$
is large.
Here we use a simple and generic enumerative program synthesis algorithm.
%% which we can accomplish by adding new programs to the frontiers means searching for new programs $p$ for task $x$
%% where  is large.
\\\noindent \textbf{Sleep-G: Maxing $ \lowerBound$ w.r.t.\ the DSL.} Here $\left\{\mathcal{F}_x \right\}_{x\in X}$ is held fixed, and so we can evaluate $\lowerBound$. Now the problem is that of searching the discrete space of DSLs and finding one maximizing $\mathcal{D}$.
To represent DSLs and search the space of DSLs, we use Fragment Grammars~\cite{tim}.

Searching for programs is hard because
of the large combinatorial search space. We ease this difficulty by training a neural recognition model, $q(\cdot |\cdot )$,
during the \textbf{Sleep-R} phase: $q$ is trained to approximate the
posterior over programs, $q(p|x)\approx \probability[p|x,\mathcal{D}]\propto\probability[x|p]\probability[p|\mathcal{D}}]$,
  thus amortizing the cost of finding programs with high posterior probability.

\textbf{Sleep-R: tractably maxing $\lowerBound$ w.r.t. the
  frontiers.}  Here we train %% a neural network, $q$, to predict a
%% distribution over programs conditioned on a task. The objective of $q$
%% is
$q(p|x)$ to assign high probability to programs $p$ where
$\probability[x|p]\probability[p|\mathcal{D}]$ is large, because including those programs
in the frontiers will most increase $\lowerBound$.
Concretely, the loss for $q(\cdot |x)$ is:
$$
\expect_X\left[ \text{KL}\left(\probability[\cdot |x,\mathcal{D}]\; ||\; q(\cdot |x) \right) \right]
$$
which minimizes the KL between the distribution predicted by $q$ and the true posterior over programs solving task $x$.

>>>>

The FlashFill system~\cite{gulwani2011automating} embedded in
Microsoft Excel solves problems such as these and is probably the best
known practical program-induction algorithm, but researchers in
programming languages and AI have had successes in many domains, such as handwriting recognition and
generation~\cite{lake2015human}, procedural
graphics~\cite{ellis2017learning,DBLP:journals/corr/abs-1804-01118}, question
answering~\cite{johnson2017clevr} and robot motion
planning~\cite{devlin2017neural}.  These systems
work in different ways, but most hinge upon a carefully engineered
\textbf{Domain Specific Language (DSL)}.  This is especially true for
systems like FlashFill that induce a wide range of programs
very quickly, in a few seconds or less.  DSLs constrain the search
over programs with strong prior knowledge in the form of a restricted
inventory of programming primitives finely tuned to the domain: for
text editing, these are operations like appending and
splitting on characters.

In this work, we consider the problem of building agents that learn to solve
program induction tasks, and also the problem of acquiring the prior knowledge
necessary to quickly solve these tasks in a new domain.  Representative problems
in three domains are shown in Table~\ref{initialExampleDSL}.  Our solution is an
algorithm that grows or boostraps a DSL while jointly training a neural network
to help write programs in the increasingly rich DSL.

\begin{table*} %[t!]
  \makebox[\textwidth][c]{
    \scriptsize
  \tabcolsep=4pt
  \renewcommand\code\texttt
  \renewcommand\codechar[1]{\texttt{"#1"}}
  \newcommand{\helpSize}{0.25cm}
  \begin{tabular}{>{\hspace{-0em}}c<{\hspace{-1em}}>{\hspace{-1em}}c<{\hspace{-1em}}>{\hspace{-2.5em}}c<{\hspace{-0.8em}}>{\hspace{-1em}}c<{\hspace{-1em}}}
    \toprule
    &{\normalsize List Functions}&{\normalsize Text Editing}&{\normalsize Symbolic Regression}\\\midrule
    \rotatebox[origin=c]{90}{\normalsize \pop{Programs} \& Tasks}&{\tabcolsep=7pt
      \begin{tabular}{cc}
        \begin{tabular}{c}
          \code{[7\, 2\, 3]}$\to $\code{[7\, 3]}         \\
          \code{[1\, 2\, 3\, 4]}$\to $\code{[3\, 4]} \\
          \code{[4\, 3\, 2\, 1]}$\to $\code{[4\, 3]} \\
          \pop{\code{$f(\ell) = $}\code{($f_1$ $\ell$ ($\lambda$ (x)}}\\
          \hspace{1.15cm}\pop{\code{(> x 2)))}}       \\
          \\
          \code{[2\, 7\, 8\, 1]}$\to $\code{8}               \\
          \hspace{0.15cm}\code{[3\, 19\, 14]}$\to $\code{19}                \\
          \pop{\code{$f(\ell) = $}\code{($f_2$ $\ell$)}}
        \end{tabular}
        &
        \hspace{-0.3cm}\begin{tabular}{c}
          \code{[7\, 3]}$\to $\code{False}                              \\
          \hspace{0.3cm}\code{[3]}$\to $\code{False}                    \\
          \hspace{-0.3cm}\code{[9\, 0\, 0]}$\to $\code{True\phantom{e}} \\
          \hspace{0.3cm}\code{[0]}$\to $\code{True\phantom{e}}                        \\
          \hspace{-0.3cm}\code{[0\, 7\, 3]}$\to $\code{True\phantom{e}}                \\
          \pop{\code{$f(\ell) = $}\code{($f_3$ $\ell$ 0)}}
        \end{tabular}
      \end{tabular}
    }
    &
    \hspace{-0.5cm}\begin{tabular}{c}
      +106 769-438$\to $106.769.438\\%&Nancy FreeHafer $\longrightarrow$ Dr. Nancy\\
      +83 973-831$\to $83.973.831\\
      \pop{$f(\text{\code{s}}) = $\code{(}$f_0$\code{  \codechar{.} \codechar{-}
      }}\\
      \hspace{1.25cm}\pop{\code{($f_0$ \codechar{.} \codechar{ }}}\\
      \hspace{1.4cm}\pop{\code{(cdr s)))}}\\
      ~\\
      Temple Anna H $\to $TAH\\
      Lara Gregori$\to $LG\\
      \pop{$f(\text{\code{s}}) = $\code{(}$f_2$\code{ s)}}\\
    \end{tabular}
    &
    \begin{tabular}{cc}
      \includegraphics[width = 3em]{figures/functions/4.png}&
      \includegraphics[width = 3em]{figures/functions/146}\\
      \pop{\code{$f($x$) = $($f_1$ x)}}&    \pop{\code{$f($x$) = $($f_6$ x)}}\\
      ~\\
      \includegraphics[width = 3em]{figures/functions/112.png}&
        \includegraphics[width = 3em]{figures/functions/92.png}
      \\
      \pop{\code{$f($x$) = $($f_4$ x)}}&    \pop{\code{$f($x$) = $($f_3$ x)}}\\

@@

  \end{tabular}
  \\\bottomrule\\
\end{tabular}}\vspace{-0.5cm}
\caption{Top: Tasks from three domains we apply our algorithm to, each followed by the programs \system discovers for them. Bottom: Several examples from learned DSL. Notice that learned DSL primitives can call each other, and that \system rediscovers higher-order functions like \code{filter} ($f_1$ under List Functions)}\label{initialExampleDSL}%\vspace{-0.5cm}
\end{table*}

Because any learning problem can in principle be cast as
program induction, it is important to delimit our focus.  In contrast
to computer assisted programming~\cite{solar2008program} or genetic
programming~\cite{DBLP:books/daglib/0070933}, our goal is not to automate software
engineering, or to synthesize large bodies of code starting from scratch.  Ours is a basic AI goal:
capturing the human ability to learn to think flexibly and efficiently
in new domains --- to learn what you need to know about a domain so you
don't have to solve new problems starting from scratch.  We are
focused on problems that people solve relatively
quickly, once they acquire the relevant domain expertise.  These
correspond to tasks solved by short programs --- if you have an
expressive DSL.  %% Even with a good DSL, program search may be
%% intractable;
%
% , and adding new library routines to the DSL only broadens the
% search space;
%
%% so we train a
%% neural network to guide the search procedure.

@@

Our algorithm takes inspiration from several ways that skilled human
programmers have learned to code: Skilled coders build libraries of
reusable subroutines that are shared across related programming tasks,
and can be composed to generate increasingly complex and powerful
subroutines.  In text editing, a good library should support routines
for splitting on characters, but also specialize these routines to
split on particular characters such as spaces or commas that are
frequently used to delimit substrings across tasks.  Skilled coders
also learn to recognize what kinds of programming idioms and library
routines would be useful for solving the task at hand, even if they
cannot instantly work out the details.  In text editing, one might
learn that if outputs are consistently shorter than inputs, removing
characters is likely to be part of the solution; if every output
contains a constant substring (e.g., ``Dr.''), inserting or appending
that constant string is likely to be a subroutine.

Our algorithm is called \system because it is based on a
novel kind of ``wake-sleep'' learning (c.f. \cite{hinton1995wake}), iterating
between ``wake'' and ``sleep'' phases to achieve three goals: finding
programs that solve tasks; creating a DSL by discovering
and reusing domain-specific subroutines; and training a neural network
that efficiently guides search for programs in the DSL.
The learned DSL effectively encodes a prior on programs likely to
solve tasks in the domain, while the neural net looks at the example
input-output pairs for a specific task and produces a ``posterior''
for programs likely to solve that specific task.  The neural network
thus functions as a \textbf{recognition model} supporting a form of
approximate Bayesian program induction, jointly trained with a
\textbf{generative model} for programs encoded in the DSL, in the
spirit of the Helmholtz machine~\cite{hinton1995wake}. The
recognition model ensures that searching for programs remains
tractable even as the DSL (and hence the search space for programs)
expands.

%% Our \system
%% algorithm incorporates these insights by iterating
Concretely, our algorithm iterates through three phases. 
The \textbf{Wake} phase takes a given set of \textbf{tasks}, typically
several hundred, and searches for compact programs that solve these tasks,
guided by the current DSL and neural network.
The \textbf{Sleep-G} phase grows the DSL (or \textbf{G}enerative model),
which allows the agent to more compactly
write programs in the domain. We modify the structure of the DSL by
discovering regularities across programs, compressing
them to distill out common abstractions across  programs discovered during waking.\footnote{This is loosely biologically
    inspired by the formation of abstractions during sleep memory
    consolidation~\cite{DUDAI201520}}
The \textbf{Sleep-R} phase improves the search procedure by training a neural network (the \textbf{R}ecognition model) to
write programs in the current DSL, in the spirit of ``amortized'' or
``compiled'' inference~\cite{le2016inference}.
We train the recognition model on  two data sources: samples
from the DSL (as in the Helmholtz Machine's ``sleep'' phase), and
programs found during waking.%%  As the DSL
%% improves and as search finds more programs, the recognition model gets
%% both more data to train on, and better data.
\footnote{These two sources are also
    loosely biologically inspired by the distinct episodic replay and
    hallucination components of dream sleep~\cite{fosse2003dreaming}}

\textbf{Brisk overview of related work.} Prior work on program learning has largely assumed a fixed, hand-engineered DSL,
both in classic symbolic program learning approaches (e.g., Metagol:~\cite{muggleton2015meta},
FlashFill:~\cite{gulwani2011automating}),
neural approaches  (e.g., RobustFill:~\cite{devlin2017robustfill}), and hybrids of neural and
symbolic methods (e.g., Neural-guided deductive search:~\cite{ngds}, DeepCoder:~\cite{balog2016deepcoder}).
A notable exception is the EC algorithm~\cite{Dechter:2013:BLV:2540128.2540316},
which also learns a library of subroutines.
We find EC motivating, and go beyond it and other prior work through the following contributions: %% \begin{itemize}
%%   \setlength\itemsep{0}
%%   \item We show how to learn-to-learn programs in an expressive Lisp-like programming language;
%%   \item We give an algorithm for learning DSLs, built on a formalism known
%%     as Fragment Grammars~\cite{tim}; and 
%%   \item We give a hierarchical Bayesian framing of the problem that allows a
%%     joint inference of the DSL and recognition model.
%% \end{itemize}
\\\textbf{Contributions.} (1) We show how to learn-to-learn programs in an expressive Lisp-like programming language, including conditionals, variables, and higher-order recursive functions;
(2) We give an algorithm for learning DSLs,
built on a formalism known as Fragment Grammars~\cite{tim};
and (3) We give a hierarchical Bayesian framing enabling joint inference of the DSL and recognition model.
up

@@

 We first mathematically describe our 3-step algorithm as
 an inference procedure for a hierarchical Bayesian model (Section~\ref{mathematicalFraming}),
  and then describe each step algorithmically in detail (Section~\ref{explorationSection}-\ref{grammarInductionSection}).

 \subsection{Hierarchical Bayesian Framing}\label{mathematicalFraming}

\system takes as input a set of \textbf{tasks}, written $X$, each of which is a program synthesis problem.
It has at its disposal a domain-specific \emph{likelihood model}, written $\probability[x|p]$, which scores the likelihood of a task $x\in X$ given a program $p$.\footnote{For example, for string editing,
  the likelihood is 1 if the program predicts the observed outputs on the observed inputs,
  and 0 otherwise; when learning a generative model or probabilistic program, the likelihood is
the probability of the program sampling the observation.}
Its goal is to solve each of the tasks by writing a program,
and also to infer a DSL, written $\mathcal{D}$.
We equip $\mathcal{D}$ with a real-valued weight vector $\theta$, and together
$(\mathcal{D},\theta)$ define a generative model over programs.
We frame our goal as maximum a posteriori (MAP) inference of $(\mathcal{D},\theta)$ given $X$.
Writing $J$ for the joint probability of $(\mathcal{D},\theta)$ and $X$, we want the $\mathcal{D}^*$ and $\theta^*$ solving:
\begin{align}\label{intractableObjectives}
\nonumber  J(\mathcal{D},\theta)\triangleq \probability[\mathcal{D},\theta]&\prod_{x\in X} \sum_p \probability[x|p]\probability[p|\mathcal{D},\theta]\\
  \mathcal{D}^* = \argmax_{\mathcal{D}}\int J(\mathcal{D},\theta)\;\mathrm{d}\theta &\qquad
  \theta^* =\argmax_\theta J(\mathcal{D}^*,\theta)
\end{align}

%% Inference in this model is
%% difficult because the programs are unobserved,
%% and so we must solve a hard search problem to recover them. To make
%% search tractable we learn a bottom-up \emph{recognition
%%   model} (written $q(\cdot )$, Fig.~\ref{graphicalModel}(b)).
%% The recognition model $q(\cdot )$ is a neural
%% network that regresses from input/output pairs to a distribution over
%% programs likely to explain the input/outputs. We can also view $q(\cdot )$ as implementing an amortized
%% inference
%% scheme~\cite{le2016inference}.
%%  The neural recognition model and the
%% generative model embodied in the DSL jointly train each other, as they
%% iteratively learn to solve more programming tasks.

The above equations summarize the problem from the point of view of an ideal Bayesian learner.
However, Eq.~\ref{intractableObjectives}
is wildly intractable because evaluating $J(\mathcal{D},\theta)$ involves
summing over the  infinite set of all programs.
In practice we will only ever be able to sum over a finite set of programs.
So, for each task, we define a finite set of programs, called a \emph{frontier}, and only marginalize over the frontiers:
\\\noindent\textbf{Definition.} A \emph{frontier of task $x$}, written $\mathcal{F}_x$,
is a finite set of programs s.t. $\probability[x|p] > 0$ for all $p\in \mathcal{F}_x$.

Using the frontiers we  define the following intuitive lower bound on the joint probability, called $\lowerBound$:
\begin{align}
 J\geq \lowerBound\triangleq\probability[\mathcal{D},\theta]\prod_{x\in X} \sum_{p\in \mathcal{F}_x} \probability[x|p]\probability[p|\mathcal{D},\theta]
\end{align}
We alternate maximization of $\lowerBound$ w.r.t. $\left\{\mathcal{F}_x \right\}_{x\in X}$  (\textbf{Wake})
and $(\mathcal{D},\theta)$ (\textbf{Sleep-G}):

\\\noindent \textbf{Wake: Maxing $\lowerBound$ w.r.t.\ the frontiers.} Here $(\mathcal{D},\theta)$ is fixed and we
want to find new programs to add to  the frontiers so that $\lowerBound$ increases the most.
$\lowerBound$ most increases by finding programs where $\probability[x|p]\probability[p|\mathcal{D},\theta]\propto\probability[p|x,\mathcal{D},\theta]$ 
is large (i.e., programs with high posterior probability).
%Here we use a simple and generic enumerative program synthesis algorithm.
%% which we can accomplish by adding new programs to the frontiers means searching for new programs $p$ for task $x$
%% where  is large.
\\\noindent \textbf{Sleep-G: Maxing $ \lowerBound$ w.r.t.\ the DSL.} Here $\left\{\mathcal{F}_x \right\}_{x\in X}$ is held fixed, and so we can evaluate $\lowerBound$. Now the problem is that of searching the discrete space of DSLs and finding one maximizing $\int \lowerBound \;\mathrm{d}\theta$,
and then updating $\theta$ to $\argmax_\theta \lowerBound(\mathcal{D},\theta,\left\{\mathcal{F}_x \right\})$.
%To represent DSLs and search the space of DSLs, we use Fragment Grammars~\cite{tim}.

Searching for programs is hard because
of the large combinatorial search space. We ease this difficulty by training a neural recognition model, $q(\cdot |\cdot )$,
during the \textbf{Sleep-R} phase: $q$ is trained to approximate the
posterior over programs, $q(p|x)\approx \probability[p|x,\mathcal{D}]\propto\probability[x|p]\probability[p|\mathcal{D}}]$.
  Thus training the neural network amortizes the cost of finding programs with high posterior probability.

\textbf{Sleep-R: tractably maxing $\lowerBound$ w.r.t. the
  frontiers.}  Here we train %% a neural network, $q$, to predict a
%% distribution over programs conditioned on a task. The objective of $q$
%% is
$q(p|x)$ to assign high probability to programs $p$ where
$\probability[x|p]\probability[p|\mathcal{D},\theta]$ is large, because incorporating those programs
into the frontiers will most increase $\lowerBound$.

Intuitively, this 3-phase inference procedure can work in practice because each of the 3 phases bootstraps off of the others (Figure~\ref{feeding}). As the DSL grows and as the recognition model becomes more accurate, waking becomes more effective, allowing the agent to solve more tasks; when we solve more tasks during waking,
the Sleep-G phase has more data from which to learn the DSL; and, because Sleep-R is trained on both
samples from the DSL and programs found during waking,
the recognition model gets both more data, and higher-quality data, whenever the DSL improves and whenever we discover more successful programs.
\begin{figure}\centering
  \begin{tikzpicture}
    \begin{scope}[shift = {(1,-1)}]
    \node[align = center](synthesis) at (6,4) {Search for \\programs: $p$};
    \node[align = center](DSL) at (3,1) {DSL: $\mathcal{D}$};
    \node[align = center](recognitionModel) at (9,1) {Recognition \\model: $q$};

    \draw [->,thick] (synthesis.-120) to[out = -150,in = 60] node[below,rotate = 45,align = center]{{\footnotesize Trains}\\{\footnotesize (Sleep-G)}} (DSL.30);
    \draw [->,thick] (synthesis.-60) to[out = -30,in = 120] node[below,rotate=-45,align = center]{{\footnotesize Trains}\\{\footnotesize (Sleep-R)}} (recognitionModel.150);
    \draw [->,thick] (DSL.east) to[out = -30,in = 210] node[above, align = center]{{\footnotesize Trains}\\{\footnotesize (Sleep-R)}} (recognitionModel.west);

    \draw [->,thick,dashed] (DSL.north) to[out = 90,in = 180] node[fill=white,align = center]{  \footnotesize{Inductive bias}\\\footnotesize{(Wake)}} (synthesis.west);
    \draw [->,thick,dashed] (recognitionModel.north) to[out = 90,in = 0] node[fill=white,align = center]{{\footnotesize Makes tractable}\\{\footnotesize (Wake)}} (synthesis.east);
  \end{scope}
    \end{tikzpicture}
  \caption{\system solves for programs, the DSL, and a recognition model. Each of these steps bootstrap off of the others in a Helmholtz-machine inspired wake/sleep inference algorithm.}  \label{feeding}
\end{figure}
\begin{figure}[h]
\centering
%  \includegraphics{figures/iterations.eps}
\begin{tikzpicture}
  \begin{scope}[shift={(2.5,0)}]
    \node[obs] at (3,3) (dx){$\mathcal{D}$};
    \node[latent] at (3.5,1.75) (zp){$p$};
%    \node[obs] at (4,3) (tx){$\theta^{(x)}$};
    \node[obs] at (3.5,0.7) (xp) {$x$};
    \node[align = center] at ([yshift = -0.6cm,xshift = 0.5cm]xp.south) {Wake: Infer $p$};
    \draw [->,red] (xp.east) to[out = 0,in = 90] node(nn){} (zp.north);
%    \draw [->,red] (tx) -- (zp);
    \draw [->] (dx) -- (zp);
    \node at (nn) {
      \begin{tikzpicture}[x=2.5cm,y=1.25cm,transform canvas={scale=0.2,shift={+(-1,2.5)}}]
        \tikzstyle{neuron}=[circle,fill=blue!50,minimum size=20pt]
        \fill[fill=white] (-0.25,-0.5) rectangle (2.25,-4.5);
        \node[rectangle] at (1,1) {};
        \foreach \name / \y in {1,...,4}
            \node[neuron] (I-\name) at (0,-\y) {};
        \foreach \name / \y in {1,...,3}
            \node[neuron] (H-\name) at (1,-\y-0.5) {};
        \foreach \name / \y in {1,...,4}
            \node[neuron] (O-\name) at (2,-\y) {};
        \foreach \source in {1,...,4}
            \foreach \dest in {1,...,3}
                \draw [-latex] (I-\source) -- (H-\dest);
        \foreach \source in {1,...,3}
            \foreach \dest in {1,...,4}
                \draw [-latex] (H-\source) -- (O-\dest);
      \end{tikzpicture}
    };
    \node[shift={+(0,-0.65)}] at (nn) [fill=white]{ $q(x)$ };
  \end{scope}
